---
title: "dada2 with OG fastq files"
author: "Chris"
date: "2024-09"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load packages
```{r}

#if(!requireNamespace("BiocManager")){
#  install.packages("BiocManager")
#}
#BiocManager::install("phyloseq")

```

```{r}

#if (!requireNamespace("BiocManager", quietly=TRUE))
#    install.packages("BiocManager")
#BiocManager::install("dada2", version= "3.18")

#packageVersion("dada2")


```

# load packages
```{r}

library("phyloseq")
library("dada2")

```

# read in sequences
```{r}

getwd()

path <- "/Users/cmantegna/Desktop/ysequences" 
fnFs <- sort(list.files(path, pattern = "_R1_001.fastq.gz", full.names = TRUE))
fnRs <- sort(list.files(path, pattern = "_R2_001.fastq.gz", full.names = TRUE))

```

# find the missing reverse read - only needed to do once. Sample: 5t 125 is missing. Removed from the data set to proceed
```{r}

# get sample names for each by removing the ending
#sampleNamesFs <- gsub("_R1_001.fastq.gz", "", basename(fnFs))
#sampleNamesRs <- gsub("_R2_001.fastq.gz", "", basename(fnRs))

# find the missing one
#missingReverse <- setdiff(sampleNamesFs, sampleNamesRs)

# result
#print(missingReverse)

```

# verify filtering parameters, current left several sequences out
```{r}

# remove the trunLen command because several reverse reads were left to 0. (, truncLen=c(240,160)) 

filtFs <- file.path(path, "filtered", basename(fnFs))
filtRs <- file.path(path, "filtered", basename(fnRs))
out <- filterAndTrim(fnFs, filtFs, fnRs, filtRs,
                     maxN=0, maxEE=c(3,5), truncQ=2, rm.phix=TRUE,
                     compress=TRUE, multithread=TRUE, verbose= TRUE)

```


# review output
```{r}
# 5 sequence pairs that returned 0: 1aa - 130, 1x - 119, 5d - 90, 5dd - 135, 5gg - 138
# pairs well below other sequences: 5J - 107 (70, 59), 5w - 128 (13, 13)
print(out)

# write to table
write.csv(out,"/Users/cmantegna/Documents/YellowIsland2023/filterOutput.csv", row.names = FALSE)

```

# remove empties
```{r}

nonEmptySamples <- out[,2] > 0  

fnFs_filtered <- filtFs[nonEmptySamples]
fnRs_filtered <- filtRs[nonEmptySamples]

```



# error rates
```{r}

errF <- learnErrors(fnFs_filtered, multithread=TRUE)
errR <- learnErrors(fnRs_filtered, multithread=TRUE)

```

# plot error rates
```{r}
plotErrors(errF, nominalQ=TRUE)

```

```{r}

plotErrors(errF, nominalQ=TRUE)

```

# alternate 1 - remove trunc and expand maxEE (mismatch adjustment)
```{r}

out1 <- filterAndTrim(fnFs, filtFs, fnRs, filtRs,
                     maxN=0, maxEE=c(5,7), truncQ=2, rm.phix=TRUE,
                     compress=TRUE, multithread=TRUE, verbose= TRUE)

```

# alternate 2 - remove trunc and expand trunQ (quality score adjustment)
```{r}

out2 <- filterAndTrim(fnFs, filtFs, fnRs, filtRs,
                     maxN=0, maxEE=c(5,7), truncQ=1, rm.phix=TRUE,
                     compress=TRUE, multithread=TRUE, verbose= TRUE)

```

# remove duplicates
```{r}

derepFs <- derepFastq(fnFs_filtered)
derepRs <- derepFastq(fnRs_filtered)

# the derep process takes a ton of memory - I was able to work through the F reads, but not the R. Trying this to continue to move forward.
# Initialize an empty list to store dereplicated reverse reads
derepRs <- vector("list", length(fnRs_filtered))

# Loop over the reverse reads and dereplicate
for (i in seq_along(fnRs_filtered)) {
  derepRs[[i]] <- derepFastq(fnRs_filtered[[i]])
}

# Check the output to confirm
derepRs


```

# move large files from memory to free up space to move forward - come back to add reverse files when finished
```{r}
# save- forward 
saveRDS(dadaFs, "dadaFs.rds")
saveRDS(derepFs, "derepFs.rds")
saveRDS(errF, "errF.rds")
saveRDS(filtFs, "filtFs.rds")
saveRDS(fnFs, "fnFs.rds")
saveRDS(fnFs_filtered, "fnFs_filtered.rds")

# save- reverse 
saveRDS(dadaRs, "dadaRs.rds")
saveRDS(derepRs, "derepRs.rds")
saveRDS(errR, "errR.rds")
saveRDS(filtRs, "filtRs.rds")
saveRDS(fnRs, "fnRs.rds")
saveRDS(fnRs_filtered, "fnRs_filtered.rds")

# Remove from memory
rm(dadaFs)
rm(derepFs)
rm(errF)
rm(filtFs)
rm(fnFs)
rm(fnFs_filtered)
gc()

# Reload (when needed)
#dadaFs <- readRDS("dadaFs.rds")

```


# infer asv
```{r}

dadaFs <- dada(derepFs, err=errF, multithread=TRUE)

dadaRs <- dada(derepRs, err=errR, multithread=TRUE)


```

# reload Fs and merge files
```{r}

# reload 
dadaFs <- readRDS("dadaFs.rds")
derepFs<- readRDS("derepFs.rds")

# merge
mergers <- mergePairs(dadaFs, derepFs, dadaRs, derepRs, verbose=TRUE)


```

