---
title: "LE Yellow Island Visualizations"
author: "Chris"
date: "2023-07-31"
output: html_document
output: 
  html_document: 
  theme: readable 
  highlight: zenburn 
    toc: true 
    toc_float: true 
    number_sections: true 
    code_folding: show 
    code_download: true 
editor: 
markdown: 
wrap: 72 
---

###READ ME We will use the following files to explore our data

algae.csv - the algal presence/ absence by top layer and second layer percentCover.csv - the substrate and second layer information 
truncated.csv - this is a simplified version of species data
combinedMarine.csv - this is the post point and manchester data from MARINe

# Package Install and call
```{r}

# Installation is commented out, remove hashtag if you need to install
#install.packages("kableExtra")
#install.packages('gridExtra')

```

```{r}

#Use library to call your installed packages for use
library(data.table)
library(dplyr)
library('ggplot2')
library("indicspecies")
library(kableExtra)
library(knitr)
library(RColorBrewer)
library(tidyr)
library(tidyverse)
library(vegan)

```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Know where you're working
```{r}

# make sure you know where you are working, remove the #if you need to make this work
#getwd()
#setwd()

```

# Load your data
```{r}

counts<- read.csv("truncated.csv", header= TRUE) #species counts per section
algae<- read.csv("algae.csv", header= TRUE) #algal cover in first and second layer of each quadrat
substrate<- read.csv("substrate.csv", header=TRUE) #substrate percent cover in first and second layer of each quadrat
marine<- read.csv("combinedMarine.csv", header= TRUE)

```

# Inspect your data
```{r}

# remove the hashtag from each operator to use it for each of your data sets
# check out row/ column counts to make sure you've got everything
#dim(counts)

#quick view of data frame structure to see character v factor and such
#str()

#look at the summary of your data to see what categorically exists
#summary()

```
# Big Picture Viz
# Manipulate counts data to make visualizations
```{r}

counts1 <- counts %>%
  gather(key = "Phylum", value = "Count", ANNELIDA:OSTEICHTHYES)

# Sum the counts for each SECTION and Phylum
section_phylum_counts <- counts1 %>%
  group_by(SECTION, Phylum) %>%
  summarize(Total_Count = sum(Count))

# Plot the stacked bar graph for counts by SECTION with phyla as fill
ggplot(section_phylum_counts, aes(x = SECTION, y = Total_Count, fill = Phylum)) +
  geom_bar(stat = "identity") +
  labs(x = "SECTION", y = "Count", title = "Counts for each SECTION by Phylum") +
  theme_minimal()

```

# Same Graph as above but the colors and X-axis have been cleaned up
```{r}

# Get unique Phylum values
unique_phyla <- unique(section_phylum_counts$Phylum)

# Generate a color palette with enough colors to match the number of unique Phylum values
n_colors <- length(unique_phyla)
#custom_colors <- brewer.pal(n_colors, "Spectral")
custom_colors <- c("sienna", "red", "yellow", "royalblue", "purple","seagreen", "black", "orange", "lightgrey")


plot_width <- 10
plot_height <- 6
margin_bottom <- 1.5

# Plot the stacked bar graph for counts by SECTION with phyla as fill and custom colors
ggplot(section_phylum_counts, aes(x = SECTION, y = Total_Count, fill = Phylum)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = custom_colors) +  # Set custom colors for the phyla
  labs(x = "SECTION", y = "Count") +
  ggtitle("Phyla Count by Section") +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, face = "bold"),
    plot.margin = unit(c(margin_bottom, 0.5, 0.5, 0.5), "cm"),
    plot.title = element_text(hjust = 0.5),  # Center the plot title
    plot.background = element_rect(fill = "white"),  # Set plot background color
    panel.grid = element_blank(),  # Remove grid lines
    axis.line = element_line(size = 0.5),  # Customize axis line width
    legend.position = "right"  # Move the legend to the right
  ) +
  coord_cartesian(ylim = c(0, max(section_phylum_counts$Total_Count) * 1.1))  # Adjust Y-axis limits# Plot the stacked bar graph for counts by SECTION with phyla as fill and custom colors
ggplot(section_phylum_counts, aes(x = SECTION, y = Total_Count, fill = Phylum)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = custom_colors) +  # Set custom colors for the phyla
  labs(x = "SECTION", y = "Count", title = "Phyla Count by Section") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 15, hjust = 1, vjust = 0.5, face = "bold"))


```

# Next Viz
# fix duplicate M and match the colors to the graph above
```{r}

# Sum the counts for each ZONATION and Phylum
zonation_phylum_counts <- counts1 %>%
  group_by(ZONATION, Phylum) %>%
  summarize(Total_Count = sum(Count))

# Plot the bar graph for PHYLUM counts across ZONATION
ggplot(zonation_phylum_counts, aes(x = ZONATION, y = Total_Count, fill = Phylum)) +
  geom_bar(stat = "identity") +
  labs(x = "ZONATION", y = "Phylum Count", title = "Phylum Count across ZONATION") +
  theme_minimal()

```

# MARINe Site Comparison
```{r}

# Combine Site and Section to form the X-axis label
marine$Site_Section <- paste(marine$SITE, marine$SECTION, sep = "-")

# Define the custom order of Phylum categories
custom_order <- c("Rhodophyta", "Ochrophyta", "Chlorophyta", "Mollusca", "Cnidaria",  "Arthropoda")

# Match colors to phyla from last graph
custom_colors <- c("indianred4", "salmon2", "mediumseagreen", "green2", "royalblue","red")

# Reorder the Phylum variable based on the custom order
marine$Phylum <- factor(marine$Phylum, levels = custom_order)

# Plot the stacked bar graph
ggplot(marine, aes(x = Site_Section, y = Percent_Cover, fill = Phylum)) +
  geom_bar(stat = "identity", position = "stack") +
  scale_fill_manual(values = custom_colors) +
  labs(title = "Phylum Percent Cover across Site and Section",
       x = "Site - Section",
       y = "Percent Cover") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  guides(fill = guide_legend(reverse = TRUE))  # Reverse the order of the legend

```
```{r}

```

# One more viz for the road
```{r}

```

