---
title: "LE Yellow Island Visualizations"
author: "Chris"
date: "2023-07-31"
output: html_document
output: 
  html_document: 
  theme: readable 
  highlight: zenburn 
    toc: true 
    toc_float: true 
    number_sections: true 
    code_folding: show 
    code_download: true 
editor: 
markdown: 
wrap: 72 
---

###READ ME We will use the following files to explore our data

algae.csv - the algal presence/ absence by top layer and second layer percentCover.csv - the substrate and second layer information 
truncated.csv - this is a simplified version of species data
combinedMarine.csv - this is the post point and manchester data from MARINe

# Package Install and call
```{r}

# Installation is commented out, remove hashtag if you need to install
#install.packages("kableExtra")
#install.packages('gridExtra')

```

```{r}

#Use library to call your installed packages for use
library(data.table)
library(dplyr)
library('ggplot2')
library("indicspecies")
library(kableExtra)
library(knitr)
library(RColorBrewer)
library(tidyr)
library(tidyverse)
library(vegan)

```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Know where you're working
```{r}
# make sure you know where you are working, remove the #if you need to make this work
#getwd()
#setwd()

```

# Load your data
```{r}

counts<- read.csv("truncated.csv", header= TRUE) #species counts per section
algae<- read.csv("algae.csv", header= TRUE) #algal cover in first and second layer of each quadrat
substrate<- read.csv("substrate.csv", header=TRUE) #substrate percent cover in first and second layer of each quadrat
marine<- read.csv("combinedMarine.csv", header= TRUE)
```

# Inspect your data
```{r}

# remove the hashtag from each operator to use it for each of your data sets
# check out row/ column counts to make sure you've got everything
#dim(counts)

#quick view of data frame structure to see character v factor and such
#str()

#look at the summary of your data to see what categorically exists
#summary()

```
# Big Picture Viz
# Manipulate counts data to make visualizations
```{r}

counts1 <- counts %>%
  #gather(key = "Phylum", value = "Count", ANNELIDA:OSTEICHTHYES)
```

# Next Viz
```{r}

```

# Another Viz
```{r}

```

# One more viz for the road
```{r}

```

